---
title: "Comparison of Legacy and Workflow comdat Indicator"
author: "Max Grezlik"
date: "2025-12-29"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
---

# Purpose of this comparison

Prior to 2025 (when the SOE workflow was implemented), the comdat indicator
was generated from three primary sources. Two of these data sources required
additional preprocessing steps via standalone R scripts or R Markdown files.

The historical inputs included:

- **Commercial landings**
  - `Commercial_data_pull_25.RData`
  - Derived from `comlandr::get_comland_data()`

- **Menhaden EOF**
  - `menhadenEOF.rds` (from `menhaden.Rmd`)
  - Supporting inputs:
    - `comparebait_KA.csv`
    - AM Landings by Gaichas Regions
    - maine.gov landings
    - mass.gov landings

- **NAFO landings**
  - From `get_nafo_21a_soe.R`
  - NAFO 21A and 21B

Because of challenges achieving a clean 1:1 match of all upstream inputs, this
report starts by comparing the final comdat products generated by the
legacy (ecodata) process and the new workflow.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.align = "center"
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggforce)
```

```{r load-comdat-outputs, include=FALSE}
# New workflow output

new_comdat <- readRDS("~/EDAB_Indicators/comdat.rds")

# Legacy ecodata output

ecodata_comdat <- ecodata::comdat


# Confirm identical structure

names(new_comdat)
names(ecodata_comdat)


```

Both datasets contain the same fields:

- Time
- Var
- Value
- EPU
- Units

```{r comparison-table, include=FALSE}
new_comdat2 <- new_comdat %>%
mutate(source = "new")

old_comdat2 <- ecodata_comdat %>%
mutate(source = "old")

comdat_compare <- bind_rows(new_comdat2, old_comdat2) %>%
pivot_wider(
names_from  = source,
values_from = Value
)
```

The scatterplots below show strong correlation between old and new values for most variables.
Slight differences exist for Benthivore data including MAFMC managed species.
Larger differences exist in Planktivore Seafood Landings including for NEFMC managed species (likely Herring). Herring previously was added via a static csv file. The chunk of code doing this is commented out of the script on GitHub. It is unclear if Herring data was added this way past 2022. https://github.com/NOAA-EDAB/SOE_data/blob/fca8096eb67c054cef4b98a9ca8916f5eb7860eb/R/create_comdat.R#L64-L87 

```{r scatterplots-old-vs-new}
n_per_page <- 4
n_pages <- ceiling(
n_distinct(comdat_compare$Var) / n_per_page
)

for (i in seq_len(n_pages)) {
p <- ggplot(comdat_compare, aes(x = old, y = new)) +
geom_abline(
slope = 1, intercept = 0,
linetype = "dashed", color = "gray50"
) +
geom_point(alpha = 0.6) +
facet_wrap_paginate(
~ Var,
ncol = 2,
nrow = 2,
scales = "free",
page = i
) +
labs(
title = paste(
"Old vs New comdat Values (page",
i, "of", n_pages, ")"
),
x = "Old comdat Value",
y = "New comdat Value"
) +
theme_minimal()

print(p)
}
```


```{r}
seafood <- comdat_compare |> dplyr::filter(Var %in% c("Piscivore Seafood Landings", "Planktivore Seafood Landings",
                                           "Other Seafood Landings", "Benthos Seafood Landings",
                                           "Benthivore Seafood Landings", "Apex Predator Seafood Landings") ) |> dplyr::mutate(difference = old - new)


```


Overlaying the time series shows strong correlation between data sets.
Slight differences in Revenue are due to a difference in reference year when accounting for inflation.



```{r time-series-comparison}
for (i in seq_len(n_pages)) {
p <- ggplot(comdat_compare, aes(x = Time)) +
geom_line(aes(y = old, color = "old")) +
geom_line(aes(y = new, color = "new")) +
facet_wrap_paginate(
~ Var,
ncol = 2,
nrow = 2,
scales = "free_y",
page = i
) +
labs(
title = paste(
"Old vs New comdat Time Series (page",
i, "of", n_pages, ")"
),
y = "Value",
color = "Version"
) +
theme_minimal()

print(p)
}
```

```{r quantify-differences, include=FALSE}
comdat_compare <- comdat_compare %>%
mutate(
diff = new - old,
pct_diff = ifelse(
  is.na(old) | old == 0,
  NA_real_,
  100 * (new - old) / abs(old)
)

)
```
```{r nonzero-differences}
tol <- 1e-6

comdat_compare_nonzero <- comdat_compare %>%
  filter(!is.na(diff) & abs(diff) > tol) |> 
  filter(Time <= 2023)

```




All values in new_comdat but not in ecodata_comdat come from 2024.
All values in ecodata_comdat but not in new_comdat come from EPU = OTHER

```{r coverage-differences}
anti_join(new_comdat, ecodata_comdat,
by = c("Time", "Var", "EPU"))

anti_join(ecodata_comdat, new_comdat,
by = c("Time", "Var", "EPU"))
```


Comparing input raw values for Herring

```{r}
# Raw data for new indicator
# comdat_path <- '~/EDAB_Datasets/Workflows/commercial_comdat.rds'
# comland_list <- readRDS(comdat_path)
# new_comland <- comland_list$comland |>
#     tibble::as_tibble() |>
#     dplyr::filter(NESPP3 != 221, NESPP3 != 789)

new_comdat_plank_land_GB <- readRDS("~/EDAB_Indicators/comdat.rds") |> 
                dplyr::filter(Var == "Planktivore Seafood Landings" ) |> 
                dplyr::filter(EPU == 'GB') |> 
                dplyr::select(Time, Value) |> 
                dplyr::rename(new = Value)

# Raw data for old indicator 

  load("~/EDAB_Dev/grezlik/comdat_inputs_comparison/Commercial_data_pull_25.RData")
  
  old_comland <- comdat
rm(comdat)

old_comland_plank_land_GB <- old_comland |> 
                dplyr::filter(Var ==  "Planktivore Seafood Landings" ) |> 
                dplyr::select(-Source) |> 
                dplyr::rename(EPU = Region) |> 
                dplyr::filter(EPU == 'GB') |> 
                dplyr::select(Time, Value) |> 
                dplyr::rename(old = Value)


GB_plank_land_compare <- dplyr::full_join(new_comdat_plank_land_GB, old_comland_plank_land_GB) |> 
                            dplyr::mutate(diff = old - new)

  
```









