---
title: "Comparison of Legacy and Workflow comdat Indicator"
author: "Max Grezlik"
date: "2025-12-29"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
---

# Purpose of this comparison

Prior to 2025 (when the SOE workflow was implemented), the comdat indicator
was generated from three primary sources. Two of these data sources required
additional preprocessing steps via standalone R scripts or R Markdown files.

The historical inputs included:

- **Commercial landings**
  - `Commercial_data_pull_25.RData`
  - Derived from `comlandr::get_comland_data()`

- **Menhaden EOF**
  - `menhadenEOF.rds` (from `menhaden.Rmd`)
  - Supporting inputs:
    - `comparebait_KA.csv`
    - AM Landings by Gaichas Regions
    - maine.gov landings
    - mass.gov landings

- **NAFO landings**
  - From `get_nafo_21a_soe.R`
  - NAFO 21A and 21B

Because of challenges achieving a clean 1:1 match of all upstream inputs, this
report starts by comparing the final comdat products generated by the
legacy (ecodata) process and the new workflow.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.align = "center"
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggforce)
```

```{r load-comdat-outputs, include=FALSE}
# New workflow output

new_comdat <- readRDS("~/EDAB_Indicators/comdat.rds")

# Legacy ecodata output

ecodata_comdat <- ecodata::comdat

# test output (ran workflow with old comm input)

test_comdat <- readRDS('~/EDAB_Dev/grezlik/test_indicators/comdat.rds')
# Confirm identical structure

names(new_comdat)
names(ecodata_comdat)
names(test_comdat)

```

Both datasets contain the same fields:

- Time
- Var
- Value
- EPU
- Units

```{r comparison-table, include=FALSE}
new_comdat2 <- new_comdat %>%
mutate(source = "new")

old_comdat2 <- ecodata_comdat %>%
mutate(source = "old")

test_comdat2 <- test_comdat |> 
  mutate(source = 'test')

```

The scatterplots below show strong correlation between old and new values for most variables.
Slight differences exist for Benthivore data including MAFMC managed species.
Larger differences exist in Planktivore Seafood Landings including for NEFMC managed species (likely Herring). Herring previously was added via a static csv file. The chunk of code doing this is commented out of the script on GitHub. It is unclear if Herring data was added this way past 2022. https://github.com/NOAA-EDAB/SOE_data/blob/fca8096eb67c054cef4b98a9ca8916f5eb7860eb/R/create_comdat.R#L64-L87 

```{r scatterplots-old-vs-new}
comdat_compare <- bind_rows(new_comdat2, old_comdat2) %>%
pivot_wider(
names_from  = source,
values_from = Value
)

n_per_page <- 4
n_pages <- ceiling(
n_distinct(comdat_compare$Var) / n_per_page
)

for (i in seq_len(n_pages)) {
p <- ggplot(comdat_compare, aes(x = old, y = new)) +
geom_abline(
slope = 1, intercept = 0,
linetype = "dashed", color = "gray50"
) +
geom_point(alpha = 0.6) +
facet_wrap_paginate(
~ Var,
ncol = 2,
nrow = 2,
scales = "free",
page = i
) +
labs(
title = paste(
"Old vs New comdat Values (page",
i, "of", n_pages, ")"
),
x = "Old comdat Value",
y = "New comdat Value"
) +
theme_minimal()

print(p)
}
```


```{r scatterplots-test-vs-new}
comdat_compare <- bind_rows(new_comdat2, test_comdat2) %>%
pivot_wider(
names_from  = source,
values_from = Value
)

n_per_page <- 4
n_pages <- ceiling(
n_distinct(comdat_compare$Var) / n_per_page
)

for (i in seq_len(n_pages)) {
p <- ggplot(comdat_compare, aes(x = test, y = new)) +
geom_abline(
slope = 1, intercept = 0,
linetype = "dashed", color = "gray50"
) +
geom_point(alpha = 0.6) +
facet_wrap_paginate(
~ Var,
ncol = 2,
nrow = 2,
scales = "free",
page = i
) +
labs(
title = paste(
"Test vs New comdat Values (page",
i, "of", n_pages, ")"
),
x = "Test comdat Value",
y = "New comdat Value"
) +
theme_minimal()

print(p)
}
```


```{r scatterplots-test-vs-old}
comdat_compare <- bind_rows(old_comdat2, test_comdat2) %>%
pivot_wider(
names_from  = source,
values_from = Value
)

n_per_page <- 4
n_pages <- ceiling(
n_distinct(comdat_compare$Var) / n_per_page
)

for (i in seq_len(n_pages)) {
p <- ggplot(comdat_compare, aes(x = test, y = old)) +
geom_abline(
slope = 1, intercept = 0,
linetype = "dashed", color = "gray50"
) +
geom_point(alpha = 0.6) +
facet_wrap_paginate(
~ Var,
ncol = 2,
nrow = 2,
scales = "free",
page = i
) +
labs(
title = paste(
"Test vs Old comdat Values (page",
i, "of", n_pages, ")"
),
x = "Test comdat Value",
y = "Old comdat Value"
) +
theme_minimal()

print(p)
}
```



All values in new_comdat but not in ecodata_comdat come from 2024.
All values in ecodata_comdat but not in new_comdat come from EPU = OTHER

```{r coverage-differences}
anti_join(new_comdat, ecodata_comdat,
by = c("Time", "Var", "EPU"))

anti_join(ecodata_comdat, new_comdat,
by = c("Time", "Var", "EPU"))
```


Comparing input raw values for Herring

```{r raw comm compare}
# Raw data for new indicator
comdat_path <- '~/EDAB_Datasets/Workflows/commercial_comdat.rds'
comland_list <- readRDS(comdat_path)
new_comland <- comland_list$comland |>
    tibble::as_tibble() |>
    dplyr::filter(NESPP3 != 221, NESPP3 != 789)

# Raw data for old indicator
old_comland_list <- readRDS("~/EDAB_Dev/grezlik/comdat_inputs_comparison/comlandpull_comdatbennet_wmenhaden.rds")
old_comland <- old_comland_list$comland |> 
  tibble::as_tibble() |> 
  dplyr::filter(NESPP3 != 221)

# call just herring for both
new_herring <- new_comland |> 
                  dplyr::filter(NESPP3 == 168)

old_herring <- old_comland |> 
                  dplyr::filter(NESPP3 == 168)

# old herring has more observations

# combine data sets for plotting
herring_compare_new <- new_herring |> dplyr::mutate(data_source = 'new')
herring_compare_old <- old_herring |> dplyr::mutate(data_source = 'old')

herring_compare <- dplyr::bind_rows(herring_compare_new, herring_compare_old)

# 1. Aggregate the data
herring_summary <- herring_compare |> 
  group_by(YEAR, data_source, EPU) |> 
  summarise(SPPVALUE_sum = sum(SPPVALUE, na.rm = TRUE), .groups = "drop")

# 2. Create the plot
ggplot(herring_summary, aes(x = YEAR, y = SPPVALUE_sum, color = data_source)) +
  geom_line(linewidth = 1) +
  geom_point(alpha = 0.5) + 
  facet_wrap(~EPU, scales = 'free_y') + 
  theme_minimal() +
  labs(
    title = "Total Herring Value (SPPVALUE) by Year and Source",
    subtitle = "Faceted by Ecological Production Unit (EPU)",
    x = "Year",
    y = "Total SPPVALUE",
    color = "Data Source"
  ) +
  theme(legend.position = "bottom")

# plot data sources seperately 
herring_summary |> dplyr::filter(data_source == 'new') |> 
ggplot( aes(x = YEAR, y = SPPVALUE_sum, color = data_source)) +
  geom_line(linewidth = 1) +
  geom_point(alpha = 0.5) + 
  facet_wrap(~EPU, scales = "free_y") + 
  theme_minimal() +
  labs(
    title = "Total Herring Value (SPPVALUE) by Year and Source",
    subtitle = "Faceted by Ecological Production Unit (EPU)",
    x = "Year",
    y = "Total SPPVALUE",
    color = "Data Source"
  ) +
  theme(legend.position = "bottom")

herring_summary |> dplyr::filter(data_source == 'old') |> 
ggplot( aes(x = YEAR, y = SPPVALUE_sum, color = data_source)) +
  geom_line(linewidth = 1) +
  geom_point(alpha = 0.5) + 
  facet_wrap(~EPU, scales = "free_y") + 
  theme_minimal() +
  labs(
    title = "Total Herring Value (SPPVALUE) by Year and Source",
    subtitle = "Faceted by Ecological Production Unit (EPU)",
    x = "Year",
    y = "Total SPPVALUE",
    color = "Data Source"
  ) +
  theme(legend.position = "bottom")



```













