---
title: "SOEworkflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SOEworkflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SOEworkflows)
```

To contribute an indicator workflow to this repo, please follow the steps below:

-  Clone this repo
-  Switch to `dev` branch and pull
-  Build the package Ctrl + Shift + B
-  Create branch off `dev` for your indicator. Name the branch the same name as your indicator
-  Add a main wrapper workflow function in `data-raw` folder. Name it `workflow_<ecodata_dataset_name>.r`
    * This function should take only filepaths as arguments and should save the ecodata data set (see examples that already exist in the `data-raw` folder)
    * Wrap the `workflow_<ecodata_dataset_name>` function in a `tryCatch()` statement. See existing workflows for how this is implemented
-  Add file in the `R` folder. Name this as `<ecodata_dataset_name>.r`. 
    * This file should contain the set of functions required to calculate the indicator. The format of the indicator should match the format as found in the `ecodata` package. You may need to look at the `ecodata/data-raw/get_<your_indicator>` function to see any additional formatting that is required
    * This/these functions should be called from the `workflow_<ecodata_dataset_name>.r` function
    * All functions should only take file paths as arguments. All data dependencies should be read in by passing a file path to the data.
-  Check that there is a `create_ecodata_dataset_name.Rd` file in the `man` folder. If not, run `devtools::document()` to create and commit one.
-  Add any packages used to create the indicator in the `DESCRIPTION` file.
-  Add instructions in the `tempate_readme.md` file (in the root of this repo)
-  Final testing should be done inside your Rstudio container. (You'll need to repeat the first few steps; clone, pull, build)
-  To test your workflow use the `example_test_runs.r` as a template.
-  To test the R package, please build the package and then run `devtools::check()`, either from the command line or click the `check` button under the build tab in Rstudio
   - If any errors arise (you can ignore `Notes` and `warnings`),  please try to fix them prior to creating a pull request
-  When checks pass, create a pull request into the `dev` branch and assign two people as reviewers
